Compositional Certified Resource Bounds\ifshort{\vspace{-1.15cm}}{}

### 2015-03-06 paper/poly.tex
start writing down the polynomial system

\documentclass[nocopyrightspace,preprint,pldi]{sigplanconf-pldi15}

@@

\section{Polynomial Potential}

>>>>

\documentclass[nocopyrightspace,preprint]{sigplanconf-pldi15}

@@

\section{Canonical Polynomial Potential}

The following presentation of amortized analysis for resource
inference on integer programs permits to obtain precise polynomial
bounds by naturally extending the previous linear work. Innovations
of this presentation include
\begin{enumerate}
\item the use of canonical polynoms $X^i$ at no cost on precision,
 \\ (I wish)
\item the use of negative coefficients in potential annotations,
\item the support for operations with mixed effects on multivariate potential.
\end{enumerate}

Let $V$ be a set of variables.  An \emph{index} $I \in \ind(V)$ is a
family that maps pairs of variables to non-zero natural numbers,
$$
I = (i_{(x,y)})_{(x,y) \in V^2} .
$$
%
We identify a family $I$ with the set of pairs $\{ ((x,y),i_{(x,y)})
\mid (x,y) \in V^2\}$.
Let $\ind(V)$ denote the set of all such indices.  We write $\ind$
instead of $\ind(V)$ if the set of variables $V$ is fixed or obvious
from the context.
%

The \emph{degree} of an index $I = (i_{(x,y)})$
is defined by
$
\deg(I) = \sum_{(x,y)} i_{(x,y)}.
$
We define $\ind_k(V) = \{ I \in \ind(V) \mid \deg(I) \leq k
\}$ to be the set of indices of degree at most $k$.  Remark
that with these notations, $\ind_0(V)$ for any $V$ is a
singleton---it contains only the empty index.

Let $V$ be a set of variabls.  An index $I \in \ind(V)$ denotes a
\emph{base polynomial} $P_I : \State \to \Q$ that maps a program
state $\state$ to a rational number.  The definition of $P_I$ is
$
P_I(\state) = \prod_{(x,y)} |[\state(x),\state(y)]|^{i_{(x,y)}}.
$
Remark that following the standard convention for $\prod$, the
empty index represents the constant base polynomial 1.

A \emph{resource polynomial} is a linear combination of
base polynomials with rational coefficients.  Resource polynomials
form a vector space over $\Q$.

We call \emph{potential annotations}
the coordinates of a resource polynomial in the basis of
all base polynomials.  An annotation $Q$ is readily represented
by a collection $(q_I)_I$ binding rational numbers to indices.
For example, the potential annotation $\{\{((x,y),2)\} \mapsto 0.5,
\emptyset \mapsto 3\}$ specifies the resource polynomial
$
  P_Q(\state) = 0.5\cdot|[\state(x),\state(y)]|^2 + 3.
$

The notion of degree can be generalized to potential annotations:
the annotation $Q = (q_I)_I$ has degree $\max_{q_I} \deg(I)$.
Remark that the degree of an annotation is exactly the degree of
the polynomial it represents.

\paragraph{Potential Method for Resource Analysis}

The previous potential annotations will be used to decorate
programs and bound their resource consumption.  A program
$P$ consumes $n$ resource units turning a state $\state$ into $\state'$,
we note this $\state \rightarrow^n_P \state'$.  The \emph{potential
method} make us look for two annotations $Q$ and $Q'$ such
that
$$
  \forall n\, \state\, \state'.\, \state \rightarrow^n_P \state'
    \implies P_Q(\state) \ge n + P_{Q'}(\state').
$$
The composition of these annotations through the whole program
using a quantitative logic permits to derive a resource bound.
In the sequel we explicit a relation that is sufficient to
satisfy the previous condition when the program $P$ is an
increment operation $x \gets x+y$.

\paragraph{Shifts of Resource Polynomials}

By symmetry we can consider the case where $P=x \gets x+y$ with
$y>0$, the case where $y \le 0$ is only a few signs away.  Let
$Q$ be a potential annotation, we will describe a $Q'$ and a
$P^y_.$ map such that if $\state \rightarrow_P \state'$ (with no resource cost),
$P_Q(\state) = P^y_{Q'}(\state')$.  We call $Q'$ a shift of $Q$ and
write $Q' = \qsh_x^+ Q$.  (When $y\le0$ we get $\qsh_x^-Q$.)

The shift of $Q$ is defined on an extended set of indices, namely
$\ind^\square = \ind(V \cup \{\square\})$.  This square is not a real
variable, but a helper with special semantics.
We can now pose $\qsh_x^+ Q = (q'_I)_{I\in\ind^\square}$ where,

\begin{align*}
q'_I =
q_{I[\square \gets x]}
\prod_z
   \binom{i_{(\square,z)} {+} i_{(x,z)}}{i_{(\square,z)}}
   \binom{i_{(z,\square)} {+} i_{(z,x)}}{i_{(z,\square)}}
   (-1)^{i_{(z,\square)}}
   .
\end{align*}

The substitution operation $I[\square \gets x]$ defines an index $I'$
in $\ind(V)$ with the same degree as $I$.  This index satisfies
$$
i'_{(x,z)} {=} i_{(x,z)} {+} i_{(\square,z)}, \;
i'_{(z,x)} {=} i_{(z,x)} {+} i_{(z,\square)}, \;
i'_{(u,v)} {=} i_{(u,v)} \mbox{ o.w.}
$$
Notations may seem heavy, but what they really express is that
replacing $\square$ by $x$ in $x^2\square^3z^2$ yields $x^5z^2$.

\clearpage
\section{Binomial Polynomial Potential}

This section presents the binomial version of the previous system.
In practice, negative coefficients could be enough to avoid having to use
this basis.  I also suspect that there is no significant simplifications
in the presentation using the binomial basis.
